Ensemble Stability of Biological Networks

### 2014-10-17 tex/discussion.tex
update main text

%!TEX root = ../paper.tex
% Our development suggests two predictions that could be used to test the theory described here. One is that,
Our theory predicts that any ensemble of systems where robustness has been positively selected over a sufficiently long period of time, should exhibit a bias toward more hierarchical network topologies. At the ecological level, a system subjected to the environmental stress of overfishing, which may imply selection for robustness, has been demonstrated to exhibit such a bias toward more hierarchical network architectures \cite{Bascompte2005}. In the short term, this prediction may be further evaluated at the levels of both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their robustness \cite{Zhao2006,Bhardwaj2010,Colm}.
%The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.
In the long term, this prediction may be evaluated using experimental evolution by comparing the degree of hierarchy that emerges in the evolution of gene regulatory network topology in the context of both static and fluctuating environments that differentially select for robustness \cite{Leroi1994}.

%The fact that such a relationship exists has the potential to unify our understanding of biological networks.
In order to further this work from a theoretical perspective, it will be necessary to deepen our understanding of the relationship between dynamical robustness and the underlying network topology.
%One approach is to attempt to compute robustness for systems of larger size. Using existing methods, this would require large-scale Monte Carlo simulations. However,
It may be possible to prove a bound on the robustness of individual SCCs, which could provide more definite knowledge and circumvent the need for expensive simulations. The conservation of robustness with respect to nontrivial symmetries including the interchange of SCCs and permutation within SCCs suggests the existence of an evolutionary neutral space. A deeper mathematical characterization of the full symmetry groupoid of dynamical robustness may help to characterize a fundamental evolutionary constraint \cite{Golubitsky2006}.

>>>>

%!TEX root = ../paper.tex
% Our development suggests two predictions that could be used to test the theory described here. One is that,
Our theory predicts that any ensemble of systems where robustness has been positively selected over a sufficiently long period of time should exhibit a bias toward more hierarchical network topologies. For example, at the ecological level, a system subjected to the environmental stress of overfishing, which may imply selection for robustness, has been observed to exhibit such a bias toward more hierarchical network architectures \cite{Bascompte2005}. In the short term, this prediction may be further evaluated at the levels of both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their robustness \cite{Zhao2006,Bhardwaj2010,Colm}.
%The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.
In the long term, this prediction may be evaluated using experimental evolution by comparing the degree of hierarchy that emerges in the evolution of gene regulatory network topology in the context of both static and fluctuating environments that differentially select for robustness \cite{Leroi1994}.

%The fact that such a relationship exists has the potential to unify our understanding of biological networks.
In order to further this work from a theoretical perspective, it will be necessary to deepen our understanding of the relationship between dynamical robustness and the underlying network topology.
%One approach is to attempt to compute robustness for systems of larger size. Using existing methods, this would require large-scale Monte Carlo simulations. However,
It may be possible to prove a bound on the robustness of individual SCCs, which could provide more definite knowledge and circumvent the need for expensive simulations. The conservation of robustness with respect to nontrivial symmetries including the interchange of SCCs and permutation within SCCs suggests the existence of an evolutionary neutral space. A deeper mathematical characterization of the full symmetry groupoid of dynamical robustness may thus help to characterize this potential evolutionary constraint \cite{Golubitsky2006}.

### 2014-10-17 tex/figures.tex
update main text

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.9\columnwidth]{fig/biomodelexamples.pdf}
\caption{{\bf Dynamical models in systems biology.} The top row represents a chemical reaction network (CRN) \cite{Shinar2010}, a gene regulatory network (GRN) \cite{Karlebach2008}, and an ecological regulatory network (ERN) \cite{Rohr2014} in terms of the graphical methods specific to each field mapped into the interaction graph, which provides a unified representation for networks across these fields. The second row represents a particular example of a system of differential equations that are used to model a biological network within each of the domains of application considered here \cite{Shinar2010,Karlebach2008,Shinar2010}. The third row shows the general form of a system of differential equations that can be used to model any network architecture of any size within each of the three domains of application exemplified here.}
\label{fig:biomodelexamples}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.9\columnwidth]{fig/modsccsym.pdf}
\caption{{\bf Open systems, strongly connected components and symmetries of robustness.} (A) Example of the combination of open system modules to construct closed systems. (B) strongly connected components highlighted in gray for each of the four graphs representing the interdependencies relevant to four different three component systems. The most hierarchical system in the top panel has the highest possible number of connected components, three, whereas the system in the bottom panel containing a single cycle and therefore posessing no hierarchy contains only one connected component. The two panels in the middle represent examples of hierarchical modular systems that posess both modularity and hierarchy. (C) Symmetries of the $\hier$ transformation between graphs and SCCs. The transformation a represents an interchange of SCCs, b moving a link between nodes in a component and c adding a link. All three transformations represent symmetries of the $\hier$ transformation from graphs to SCCs while only a and b are symmetries of robustness.}
\label{fig:modsccsym}
\end{figure}

@@

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.4\columnwidth]{fig/scc2.pdf}
% \caption{{\bf Example of strongly connected components.} (A) - (D) show strongly connected components highlighted in gray for each of the four graphs representing the interdependencies relevant to four different three component systems. Note that the most hierarchical system in (A) has the highest possible number of connected components, three, whereas the system containing a single cycle and therefore posessing no hierarchy contains only one connected component. Systems (B) and (C) represent examples of hierarchical modular systems that posess both modularity and hierarchy.}
% \label{fig:scc}
% \end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=1.0\columnwidth]{fig/combinedfigs.pdf}
\caption{{\bf Characterization of stability and robustness according to properties of system structure} (A) Robustness versus connectivity, (B) robustness versus hierarchy, (C) Robustness versus number of cycles and connectivity for three component systems. Each point represents the average robustness of all graphs having a given number of cycles and connectivity. (D) Robustness versus hierarchy and connectivity for three component systems. Each point represents the average robustness of all graphs having a given hierarchy and connectivity.
}
\label{fig:combined}
\end{figure}

@@

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.8\columnwidth]{fig/connectcycle3D3x3.pdf}
% \caption{{\bf Robustness versus number of cycles and connectivity for three component systems.} Each point represents the average robustness of all graphs having a given number of cycles and connectivity.}
% \label{fig:connectcycle3D3x3}
% \end{figure}

@@

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.8\columnwidth]{fig/connectdist3D3x3.pdf}
% \caption{{\bf Robustness versus hierarchy and connectivity for three component systems.} Each point represents the average robustness of all graphs having a given hierarchy and connectivity.}
% \label{fig:connectdist3D3x3}
% \end{figure}

>>>>

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.9\columnwidth]{fig/biomodelexamples.pdf}
\caption{{\bf Dynamical models in systems biology.} The top row represents a chemical reaction network (CRN) \cite{Shinar2010}, a gene regulatory network (GRN) \cite{Karlebach2008}, and an ecological regulatory network (ERN) \cite{Rohr2014} in terms of the graphical methods specific to each field mapped into the interaction graph, which provides a unified representation for networks across these fields. The second row represents a particular example of a system of differential equations that are used to model a biological network within each of the domains of application considered here. The third row shows the general form of a system of differential equations that can be used to model any network architecture within each domain.}
\label{fig:biomodelexamples}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.9\columnwidth]{fig/modsccsym.pdf}
\caption{{\bf Open systems, strongly connected components and symmetries of robustness.} (A) Example of the combination of open system modules to construct closed systems. (B) SCCs highlighted in gray for each of the four graphs representing the interdependencies relevant to four different three variable systems. The most hierarchical network, top panel, is the one that maximizes the number of SCCs and the number of links between them. We therefore define hierarchy as $max(\hbox{ED}) - \hbox{ED}$ where ED is the edit distance representing the number of link addition/deletion operations necessary to transform a given graph into the most hierarchical one. The two panels in the middle represent examples of hierarchical modular systems that posess both modularity (i.e. SCCs with more than one variable) and hierarchy. (C) Symmetries of the $\hier$ transformation between graphs and SCCs. The transformation $a$ represents an interchange of SCCs, $b$ moving a link between nodes in a component and $c$ adding a link. All three transformations represent symmetries of the $\hier$ transformation from graphs to SCCs while only $a$ and $b$ are symmetries of robustness.}
\label{fig:modsccsym}
\end{figure}

@@

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.4\columnwidth]{fig/scc2.pdf}
% \caption{{\bf Example of strongly connected components.} (A) - (D) show strongly connected components highlighted in gray for each of the four graphs representing the interdependencies relevant to four different three variable systems. Note that the most hierarchical system in (A) has the highest possible number of connected components, three, whereas the system containing a single cycle and therefore posessing no hierarchy contains only one connected component. Systems (B) and (C) represent examples of hierarchical modular systems that posess both modularity and hierarchy.}
% \label{fig:scc}
% \end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=1.0\columnwidth]{fig/combinedfigs.pdf}
\caption{{\bf Characterization of stability and robustness according to properties of system structure for three variable systems} (A) Robustness versus connectivity. The red colored points represent averages for each connectivity. The lowest and highest robustness network architectures are labelled. Other network architectures are shown in \ref{tab:structstabmat3}. (B) Robustness versus hierarchy. (C) Number of cycles and (D) hierarchy vs connectivity and robustness. The color of each point represents the average robustness of all graphs having the parameters specified on the $x$ and $y$ axes.
}
\label{fig:combined}
\end{figure}

@@

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.8\columnwidth]{fig/connectcycle3D3x3.pdf}
% \caption{{\bf Robustness versus number of cycles and connectivity for three variable systems.} Each point represents the average robustness of all graphs having a given number of cycles and connectivity.}
% \label{fig:connectcycle3D3x3}
% \end{figure}

@@

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.8\columnwidth]{fig/connectdist3D3x3.pdf}
% \caption{{\bf Robustness versus hierarchy and connectivity for three variable systems.} Each point represents the average robustness of all graphs having a given hierarchy and connectivity.}
% \label{fig:connectdist3D3x3}
% \end{figure}

### 2014-10-17 tex/introduction.tex
update main text

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014}. This implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

>>>>

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014}, which implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

### 2014-10-17 tex/supplement.tex
update main text

A strongly connected component (SCC) of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \reffigscc along with their adjacency matrices and hierarchy diagrams (to be defined below).
The decomposition of a digraph into strongly connected components corresponds to a block triangular decomposition of its adjacency matrix.  Say that the graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$, which have been labelled in such a way that there are no links from vertices in component $C_i$ to component $C_j$ when $i < j$.  Label the vertices in such a way that $v_1, \ldots, v_{n_1}$ belong to $C_1$, $v_{n_1 + 1}, \ldots, v_{n_2}$ belong to $C_2$, etc.  Then, if we choose basis vectors corresponding to this labelling of the vertices, we will have $a_{ij} = 0$ whenever $i$ and $j$ correspond to different components and $i > j$.  This condition is equivalent to stating that the matrix is block triangular with blocks of size $n_1, n_2, \ldots$.

@@

\subsection{Stability and robustness analysis of particular system ensembles}
For systems having two components, we can analytically compute the probability of stability and robustness from \ref{eq:condprobgen}. For those having three components, we can estimate these same quantities using Monte Carlo simulations. Systems of larger size can be analyzed using the symmetry properties of robustness extracted from this analysis. For two-component systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant. Suppose we have a stable matrix
$$
\begin{bmatrix}
a & b \\
d & c
\end{bmatrix}
$$
where $a + c < 0$ and $ac > bd$.  For the case in which $x_1=a,\,x_2=b,\,x_3=c,\,x_4=d$ we need to compute what corresponds to $R(\mathrm{stab},\mu'_k)$ where $k=1 \ldots 4$. By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent to resampling $d$ so we only need to explicitly compute $ R(\mathrm{stab},\mu'_1)$ and $R(\mathrm{stab},\mu'_2)$. Suppose that we resample $b$ to compute $R(\mathrm{stab},\mu'_2)$.
% \begin{strip}
% \begin{align}\label{eq:condprob}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}.
% \end{align}
The denominator of \ref{eq:condprobgen} in this case is given by
\begin{align*}
P\left(\textrm{stab } \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprobgen}
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a & b' \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a' & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

Using this approach the probability of stability and of robustness for all two component systems is given in \ref{tab:structstabmat}.

The analogous results for all three component systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \reffigrobustconnect. This process is associated with some error relative to the exact integration described above. If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

It has been stated previously on the basis of simulation that system stability decreases with connectivity as the system size goes to infinity \cite{May1972}. For small system sizes such as the two and three component systems, studied here the situation is not so clear cut. For two component systems, system stability is constant across the entire range of connectivities. For three component systems, the trend shows a minor decrease from connectivity $4$ to $5$ followed by small fluctuations as shown in \ref{fig:apstab3x3}.

The relationship between connectivity and robustness for two component systems is shown in \ref{tab:structstabmat} and likewise for three component systems in \ref{tab:structstabmat3} and \reffigrobustconnect. If we average over the different classes of matrices for a given connectivity we see there is a correlation between connectivity and robustness demonstrated by the red dots in \reffigrobustconnect.
% \subsection{Cycle number is inversely correlated with robustness}
\ref{fig:cycle3x3} shows the robustness for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}. There appears to be a weak negative correlation between robustness and the number of simple cycles.

% \subsection{Cycle number and connectivity classify robust systems}
The combination of connectivity and cycle number as shown in \reffigconnectcycle3D3x3 provides a better classification of the dependence of robustness upon network topology. Here the robustness of three component systems with a given number of cycles, increases monotonically with connectivity. The network with the highest robustness for three component systems is that of \reffigscc (top panel). This network is the most hierarchical of all three component systems in the sense that it represents a total ordering of the components of the network and its adjacency matrix also shown in \reffigscc (top panel) has a block triangular structure.

@@

The fact that the most hierarchical of networks is the most robust in the case of three component systems can be understood in general as resulting from the previously mentioned symmetry property of network robustness.  By combining our decomposition of the graph with our analysis of the robustness of random ensembles of dynamical systems, we will deduce a general result which relates robustness to hierarchy.

>>>>

A strongly connected component (SCC) of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three variable systems are outlined in \reffigscc along with their adjacency matrices and hierarchy diagrams (to be defined below).
The decomposition of a digraph into strongly connected components corresponds to a block triangular decomposition of its adjacency matrix.  Say that the graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$, which have been labelled in such a way that there are no links from vertices in component $C_i$ to component $C_j$ when $i < j$.  Label the vertices in such a way that $v_1, \ldots, v_{n_1}$ belong to $C_1$, $v_{n_1 + 1}, \ldots, v_{n_2}$ belong to $C_2$, etc.  Then, if we choose basis vectors corresponding to this labelling of the vertices, we will have $a_{ij} = 0$ whenever $i$ and $j$ correspond to different components and $i > j$.  This condition is equivalent to stating that the matrix is block triangular with blocks of size $n_1, n_2, \ldots$.

@@

\subsection{Stability and robustness analysis of particular system ensembles}
For systems having two components, we can analytically compute the probability of stability and robustness from \ref{eq:condprobgen}. For those having three components, we can estimate these same quantities using Monte Carlo simulations. Systems of larger size can be analyzed using the symmetry properties of robustness extracted from this analysis. For two-variable systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant. Suppose we have a stable matrix
$$
\begin{bmatrix}
a & b \\
d & c
\end{bmatrix}
$$
where $a + c < 0$ and $ac > bd$.  For the case in which $x_1=a,\,x_2=b,\,x_3=c,\,x_4=d$ we need to compute what corresponds to $R(\mathrm{stab},\mu'_k)$ where $k=1 \ldots 4$. By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent to resampling $d$ so we only need to explicitly compute $ R(\mathrm{stab},\mu'_1)$ and $R(\mathrm{stab},\mu'_2)$. Suppose that we resample $b$ to compute $R(\mathrm{stab},\mu'_2)$.
% \begin{strip}
% \begin{align}\label{eq:condprob}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}.
% \end{align}
The denominator of \ref{eq:condprobgen} in this case is given by
\begin{align*}
P\left(\textrm{stab } \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprobgen}
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a & b' \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a' & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

Using this approach the probability of stability and of robustness for all two variable systems is given in \ref{tab:structstabmat}.

The analogous results for all three variable systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \reffigrobustconnect. This process is associated with some error relative to the exact integration described above. If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

It has been stated previously on the basis of simulation that system stability decreases with connectivity as the system size goes to infinity \cite{May1972}. For small system sizes such as the two and three variable systems, studied here the situation is not so clear cut. For two variable systems, system stability is constant across the entire range of connectivities. For three variable systems, the trend shows a minor decrease from connectivity $4$ to $5$ followed by small fluctuations as shown in \ref{fig:apstab3x3}.

The relationship between connectivity and robustness for two variable systems is shown in \ref{tab:structstabmat} and likewise for three variable systems in \ref{tab:structstabmat3} and \reffigrobustconnect. If we average over the different classes of matrices for a given connectivity we see there is a correlation between connectivity and robustness demonstrated by the red dots in \reffigrobustconnect.
% \subsection{Cycle number is inversely correlated with robustness}
\ref{fig:cycle3x3} shows the robustness for all three variable systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}. There appears to be a weak negative correlation between robustness and the number of simple cycles.

% \subsection{Cycle number and connectivity classify robust systems}
The combination of connectivity and cycle number as shown in \reffigconnectcycle3D3x3 provides a better classification of the dependence of robustness upon network topology. Here the robustness of three variable systems with a given number of cycles, increases monotonically with connectivity. The network with the highest robustness for three variable systems is that of \reffigscc (top panel). This network is the most hierarchical of all three variable systems in the sense that it represents a total ordering of the components of the network and its adjacency matrix also shown in \reffigscc (top panel) has a block triangular structure.

@@

The fact that the most hierarchical of networks is the most robust in the case of three variable systems can be understood in general as resulting from the previously mentioned symmetry property of network robustness.  By combining our decomposition of the graph with our analysis of the robustness of random ensembles of dynamical systems, we will deduce a general result which relates robustness to hierarchy.

### 2014-10-17 tex/table2x2stab.tex
add orbit size to 3x3 table
update main text

\begin{table}[h]
\begin{tabular}{ c || c | c | c }
\hline
matrix & connectivity & \specialcell{probability of stability\\to perturbation} & probability of stability\\
\hline
  $\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}$ & 4 & 0.62 & 0.25 \\
  $\begin{pmatrix}
a & b \\
d & 0
\end{pmatrix}$, $\begin{pmatrix}
0 & b \\
d & c
\end{pmatrix}$ & 3 & 0.5 & 0.25 \\
  $\begin{pmatrix}
a & 0 \\
d & c
\end{pmatrix}$, $\begin{pmatrix}
a & b \\
0 & c
\end{pmatrix}$ & 3 & 0.67 & 0.25 \\
$\begin{pmatrix}
a & 0 \\
0 & c
\end{pmatrix}$ & 2 & 0.5 & 0.25 \\
\end{tabular}
\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for two component systems derived analytically}. All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat}
\end{table}

>>>>

\begin{table}[h]
\begin{center}
\begin{tabular}{ c || c | c | c }
\hline
matrix & connectivity & \specialcell{probability of stability\\to perturbation} & \specialcell{probability\\of stability}\\
\hline
  $\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}$ & 4 & 0.62 & 0.25 \\
  $\begin{pmatrix}
a & b \\
d & 0
\end{pmatrix}$, $\begin{pmatrix}
0 & b \\
d & c
\end{pmatrix}$ & 3 & 0.5 & 0.25 \\
  $\begin{pmatrix}
a & 0 \\
d & c
\end{pmatrix}$, $\begin{pmatrix}
a & b \\
0 & c
\end{pmatrix}$ & 3 & 0.67 & 0.25 \\
$\begin{pmatrix}
a & 0 \\
0 & c
\end{pmatrix}$ & 2 & 0.5 & 0.25 \\
\end{tabular}
\end{center}
\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for two variable systems derived analytically}. All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat}
\end{table}

### 2014-10-17 tex/tablewithgraphs.tex
reorganize
update main text

@@

\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for three component systems estimated via Monte Carlo sampling.} All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat3graph}

>>>>

%!TEX root = ../paper.tex

@@

\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for three variable systems estimated via Monte Carlo sampling.} All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat3graph}
